---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1 #

---

## Loading and preprocessing the data
(The source datafile was originally obtained from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip and is replicated here.)

**This code loads the datafile and displays some of the data:**
```{r loadcsv}
activityData = read.csv(dataFile)
head(activityData)
```
Some statistics on how much of the 'steps' data is 'NA':
```{r checkrows}
incompleteRowCount = nrow(activityData[!complete.cases(activityData), ])
message("Number of rows: ", nrow(activityData),
        "\nNumber of rows with missing date or interval values: ",
        length(which(is.na(activityData$interval) | is.na(activityData$date))),
        "\nNumber of rows with missing steps values: ",
        length(which(is.na(activityData$steps))))
```
This code determines if the NA values in the steps columns are concentrated in a few days or are spread out across all days:
```{r checkNAdistribution}
naStepsDistribution = aggregate(steps ~ date, data=activityData, function(x) {sum(is.na(x))}, na.action = NULL)
message("Number of days with partially-complete 'steps' values: ",
        nrow(naStepsDistribution[naStepsDistribution$steps > 0 & naStepsDistribution$steps < 288,]),
        "\nNumber of days with completely missing 'steps' values: ",
        nrow(naStepsDistribution[naStepsDistribution$steps == 288,]))
```
All the NA values are accounted for by 8 days for which we have no data - all other days are complete.

**The following section of code processes the data to a more useful form.**

Converting the "date" and "interval" fields to a real date value, which allows calculation of  weekday/weekend values:
```{r dateconversion}
activityData$dateTime = paste(activityData[,2],
                              floor(activityData$interval / 100),
                              activityData$interval %% 100,
                              sep="-")
activityData$dateTime = strptime(activityData$dateTime, "%Y-%m-%d-%H-%M")
```
This code marks the rows by weekday/weekend:
```{r daytype}
activityData$dayType = ifelse(weekdays(activityData$dateTime) %in% c("Saturday", "Sunday"), "weekend","weekday")
activityData$dayType = factor(activityData$dayType)
activityData$dateTime = NULL
```

## What is mean total number of steps taken per day?
This code extracts the complete rows, aggregate by date to get the daily totals. It also adds the date to the resulting table for plot generation.

```{r isolatecompleterows}
activityDataComplete = activityData[complete.cases(activityData[,1]),]
dailyStepCompleteTotals = aggregate(x = activityDataComplete$steps, by = list(activityDataComplete$date), FUN = sum)
colnames(dailyStepCompleteTotals) = c("Date", "TotalSteps")
dailyStepCompleteTotals$Date = strptime(dailyStepCompleteTotals$Date, "%Y-%m-%d")
```
**This code plots a histogram of total steps-per-day:**

```{r histogram1}
library(ggplot2)
ggplot(data=dailyStepCompleteTotals, aes(x=TotalSteps)) +
    ggtitle("Total Steps per Day (Complete Data)") +
    geom_histogram(fill="lightblue2", colour="blue", binwidth = 1000) +
    theme(plot.title = element_text(face="bold"))
```

**This code reports the mean and median values for the steps-per-day totals:**
```{r}
message("Mean steps-per-day value (complete rows): ",
        mean(dailyStepCompleteTotals$TotalSteps),
        "\nMedian steps-per-day value (complete rows): ",
        median(dailyStepCompleteTotals$TotalSteps))
```

## What is the average daily activity pattern?
Here we do a bit of extra work to calculate step averages aggregated by 5-minute interval. We attach a "weekday/weekend" factor column to the data because we'll need it a bit later.

```{r calculateStepAverages}
activityDataCompleteWeekday = activityDataComplete[activityDataComplete$dayType == "weekday",]
weekdayStepAverages = aggregate(x = activityDataCompleteWeekday$steps, by = list(activityDataCompleteWeekday$interval), FUN = mean)
weekdayStepAverages$dayType = "weekday"
activityDataCompleteWeekend = activityDataComplete[activityDataComplete$dayType == "weekend",]
weekendStepAverages = aggregate(x = activityDataCompleteWeekend$steps, by = list(activityDataCompleteWeekend$interval), FUN = mean)
weekendStepAverages$dayType = "weekend"
stepAverages = rbind(weekdayStepAverages, weekendStepAverages)
stepAverages$dayType = factor(stepAverages$dayType)
colnames(stepAverages) = c("interval", "steps", "dayType")
```

**This code makes a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):**

```{r timeseriesplot1}
library(lattice)
xyplot(steps ~ interval,
        data=stepAverages,
        main="Average Daily Activity Pattern (Complete Data)",
        type="a",
        xlab="Interval", 
        ylab="Number of Steps")
```

(As we can see, not much walking before 5:00 AM!)

**This code reports the 5-minute interval with the highest average step values:**
```{r}
message("5-minute interval with the highest step average: ", stepAverages[which.max(stepAverages$steps),"interval"])
```


## Imputing missing values
As we saw above, there are 2304 rows with missing values:
```{r}
message("Number of rows with missing data: ", incompleteRowCount)
```

There's no perfect way to deal with this.

**My strategy was to replace the NA values with the mean values for each 5-minute interval. **

Extracting this data:
```{r}
intervalMeans = aggregate(x = activityDataComplete$steps, by = list(activityDataComplete$interval), FUN = mean)
colnames(intervalMeans) = c("interval", "meanVal")
```

**Create a new dataset that is equal to the original dataset but with the missing data filled in:**

```{r}
patchedActivityData = activityData
patchedActivityData$steps = ifelse(is.na(patchedActivityData$steps), 
                            intervalMeans$meanVal[match(intervalMeans$interval, patchedActivityData$interval)],
                            patchedActivityData$steps)
```
And we aggregate the step totals:
```{r}
dailyStepTotals = aggregate(x = patchedActivityData$steps, by = list(patchedActivityData$date), FUN = sum)
colnames(dailyStepTotals) = c("Date", "TotalSteps")
```

**This code plots a histogram of total steps-per-day with the interpolated data:**
```{r histogram2}
ggplot(data=dailyStepTotals, aes(x=TotalSteps)) +
    ggtitle("Total Steps per Day (Interpolated Data)") +
    geom_histogram(fill="lightblue2", colour="blue", binwidth = 1000) +
    theme(plot.title = element_text(face="bold"))

```

Because of the interpolation strategy, noticeably more of the data is concentrated at the mean value.

**This code reports the mean and median values for the steps-per-day totals (interpolated data):**
```{r}
message("Mean steps-per-day value (with interpolated data): ",
        mean(dailyStepTotals$TotalSteps),
        "\nMedian steps-per-day value (with interpolated data): ",
        median(dailyStepTotals$TotalSteps))
```
These values are almost identical to the first mean and median. In this case, the median has become identical to the mean.

## Are there differences in activity patterns between weekdays and weekends?
**This code creates a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).:**
```{r timeseriesplot2}
xyplot(steps ~ interval | dayType,
        data=stepAverages,
        main="Average Daily Activity Pattern, Weekdays vs Weekend (Interpolated Data)",
        type="a",
        xlab="Interval", 
        ylab="Number of Steps", 
        layout=c(1,2))
```

A cursory inspection reveals some differences:

* On the weekend there isn't much walking happening before ~8 AM.
* On weekdays, there's are concentrations of activity in the morning and early evening. On weekends things are slightly more evenly distributed.

